---
title: "Time series models" 
author: "Barry Quinn"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    includes:
      after_body: ["insert-logo.html"]
    css: ["xaringan-themer.css","mycssblend.css"]
    lib_dir: libs
    nature:
      countdown: 120000
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "https://platform.twitter.com/widgets.js"
    seal: true
    self_contained: true
    footer: "Advanced Financial Data Analytics"
---

```{r setup1, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
library(fontawesome) 
library(xaringanExtra)
library(xaringanthemer)
library(tsfe)
library(fpp2)
library(tidyquant)
library(latex2exp)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
  )
```
class: middle

# Learning outcomes

.salt[
- Stationary and differencing
- Modelling stationary time series
- ARIMA models (liner model)
- Categories of algorithms in capital markets
- Algorithmic ARIMA modelling
- Practical example with some of the project data

]

---

class: middle

# Stationarity and differencing
.acid[
* The foundation of statistical inference in time series analysis is the concept of weak stationarity. ] 
.hand[A stationary series:
*  roughly horizontal
*  constant variance
*  no patterns predictable in the long-term
]
---
class: middle

.your-turn[

Are these financial time series stationary?
.pull-left[
```{r}
tsfe::indices %>%
  select(date,`RUSSELL 2000 - PRICE INDEX`) %>%
  rename(r2000=`RUSSELL 2000 - PRICE INDEX`) %>%
  drop_na() %>%
  tq_transmute(select =r2000,mutate_fun = periodReturn,type='log') ->monthly_r2002r
  ts(monthly_r2002r$monthly.returns, start = c(1988,1))->r2000r_m_ts
  autoplot(r2000r_m_ts) + ylab("Log returns") + xlab("Year") + labs(title="Figure 2: Monthly log returns of the Russell 2000 Price Index",subtitle =" from March 1988 to December 2019")
```
]
.pull-right[
```{r}
autoplot(tsfe::carnival_eps_ts) + xlab("Year") + ylab("Earnings") +
  labs(title="Figure 3:",subtitle =  "Quarterly earnings per share for Carnival Plc from the first quarter of 1994 to the fourth quarter of 2019")
```
]
]

---
class: middle 

.pull-left[
#### Inference and stationarity

* The monthly log returns of Russell 2000 index vary around zero over time.
  * If we divide up the data into subperiods we would expect each sample mean to be roughly zero.
* Furthermore, expect the recent financial crisis (2007-2009), the log returns range is approximately [-0.2,0.2].
* Statistically, the mean and the variance are constant over time OR time invariant. 
* Put together these to time invariant properties characterise a weakly stationary series.
]
.pull-right[
#### Weak stationarity and prediction

* Weak form stationarity provides a basic framework for prediction.
* For the monthly log returns of the Russell 2000 we can predict with reasonable confidence:
* Future monthly returns $\approx0$ and vary $[-0.2,0.2]$
]

---
class: middle

## Inference and nonstationarity

* Consider quarterly earnings for Carnival Plc.
* If the timespan is divided into subperiods the sample mean and variance for each period show increasing pattern.
* Earnings are **not** weakly stationary.
* There does exist models and methods for modelling such nonstationary series.

---
class: middle

.your-turn[
Is the VIX time series Stationary?

```{r}
tq_get('^VIX') %>%
  ggplot(aes(x=date,y=adjusted)) + 
  ylab("VIX") + geom_line()
```
]
---
class: middle

## Non-stationarity in the mean

.hand[Identifying non-stationary series]

.large[
* time plot.
* The ACF of stationary data drops to zero relatively quickly
* The ACF of non-stationary data decreases slowly.
* For non-stationary data, the value of $r_1$ is often
     large and positive.
]

---
class: middle

## Example: FTSE index

.panelset[
.panel[
.panel-name[time plot]
```{r}
autoplot(ftse_m_ts) + ylab("Monthly Price Index") + xlab("Year")
```
]

.panel[
.panel-name[ACF]
```{r}
ggAcf(ftse_m_ts)
```
]

.panel[
.panel-name[First differencing]
```{r}
autoplot(diff(ftse_m_ts)) + ylab("Change in monthly FTSE Index") + xlab("Year")
```
]
.panel[
.panel-name[ACF after first differencing]
```{r}
ggAcf(diff(ftse_m_ts))
```
]
]

---
class: middle

# Differencing
.fat[
* Differencing helps to **stabilize the mean**.
* The differenced series is the *change* between each observation in the original series: ${y'_t = y_t - y_{t-1}}$.
* The differenced series will have only $T-1$ values since it is not possible to calculate a difference $y_1'$ for the first observation.
]
---
class: middle

## carnival earnings ending 2010 Q1
.panelset[
.panel[
.panel-name[subset using `window()`]
```{r carnival1}
window(carnival_eps_ts,end=c(2010,1)) %>% autoplot()
```
]
.panel[
.panel-name[use `log()` to stablise variation]
```{r carnival2} 
window(carnival_eps_ts,end=c(2010,1)) %>% log() %>% autoplot()
```
]
.panel[
.panel-name[**then** seasonally difference]
```{r carnival3}
window(carnival_eps_ts,end=c(2010,1)) %>% 
  log() %>% diff(lag=4) %>%  autoplot()
```
]
.panel[
.panel-name[**then** first difference]
.hand[Is the series stationary?]

```{r carnival4}
window(carnival_eps_ts,end=c(2010,1)) %>% log() %>% diff(lag=4) %>% diff(lag=1) %>% autoplot()
```

]
]

---
class: middle

.discussion[

* Seasonally differenced series is closer to being stationary.
* Remaining non-stationarity can be removed with further first difference.
* If $y'_t = y_t - y_{t-12}$ denotes seasonally differenced series, then twice-differenced series i
* When both seasonal and first differences are applied
* it makes no difference which is done first the result will be the same.
* If seasonality is strong, we recommend that seasonal differencing be done first because sometimes the resulting series will be stationary and there will be no need for further first difference.
* It is important that if differencing is used, the differences are
interpretable.
]

---
class: middle

## Interpretation of differencing
.heat[
* first differences are the change between **one observation and the next**;
* seasonal differences are the change between **one year to the next**.

* But taking lag 3 differences for yearly data, for example, results in a model which cannot be sensibly interpreted.
]
---
class: middle

## Unit root tests
.pull-left[
>Statistical tests to determine the required order of differencing

1. Augmented Dickey Fuller test: null hypothesis is that the data are non-stationary and non-seasonal.
2. Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test: null  hypothesis is that the data are stationary and non-seasonal.
3. Other tests available for seasonal data.
]
.pull-right[

## KPSS test
```{r, echo=TRUE}
library(urca)
summary(ur.kpss(ftse_m_ts))
```

]
---
class: middle

## Automatically selecting differences

Seasonal strength $F_s = \max\big(0, 1-\frac{\text{Var}(R_t)}{\text{Var}(S_t+R_t)}\big)$

If $F_s > 0.64$, do one seasonal difference.

```{r, echo=TRUE}
carnival_eps_ts %>% log() %>% nsdiffs()
carnival_eps_ts %>% log() %>% diff(lag=4) %>% ndiffs()
```


---
class: middle

## Non-seasonal ARIMA models

#### Autoregressive models

* When $y_t$ has a statistically significant lag-1 autocorrelation, the lagged value $y_{t-1}$ might be a useful in predicting $y_t$.
* AR(1) model

$$y_{t}= c+\phi_{1}y_{t - 1} + \varepsilon_{t}$$

* where $\varepsilon_t$ is white noise.  
* This is a simple linear regression with **lagged values** of $y_t$ as predictors.

* This simple model is widely used in stochastic volatility when $y_t$ is replaced by its log volatility.

---
class: middle

## Autoregressive models

* More generally, if the  $E(y_{t-1})$ is determined by more than lag-1 we can generalise a AR(1) to an AR(p) model.

.blockquote[
Autoregressive (AR) models:
$$y_{t}= c+\phi_{1}y_{t - 1}+\phi_{2}y_{t - 2} + \dots +\phi_{p}y_{t - p}  + \varepsilon_{t},$$
where $\varepsilon_t$ is white noise.  This is a multiple linear regression with **lagged values** of $y_t$ as predictors.
]

---
class: middle

# Example of an AR(1) model

.panelset[
.panel[
.panel-name[Simulating an AR(1)]

.pull-left[
* Simulating an $y_{t} =2 -0.8 y_{t - 1}+\varepsilon_{t}$
* where $\varepsilon_t\sim N(0,1)$ for  $T=100$.

```{r arp, eval=FALSE}
set.seed(1)
autoplot(10 + arima.sim(list(ar = -0.8), n = 100)) +
  ylab("") + ggtitle("AR(1)")
```
]
.pull-right[
```{r arp1, echo=FALSE, fig.height=2}
set.seed(1)
autoplot(10 + arima.sim(list(ar = -0.8), n = 100)) +
  ylab("") + ggtitle("AR(1)")

```
]
]

.panel[
.panel-name[Simulating an AR(2)]

.pull-left[
* Simulating an $y_t = 8 + 1.3y_{t-1} - 0.7 y_{t-2} + \varepsilon_t$
* where $\varepsilon_t\sim N(0,1)$ for  $T=100$.

```{r arp3, eval=FALSE}
set.seed(1)
autoplot(20 + arima.sim(list(ar = c(1.3, -0.7)), n = 100)) + ylab("") + ggtitle("AR(2)")
```
]
.pull-right[
```{r arp4, echo=FALSE, fig.height=2}
set.seed(1)
autoplot(20 + arima.sim(list(ar = c(1.3, -0.7)), n = 100)) + ylab("") + ggtitle("AR(2)")

```
]
]
.panel[
.panel-name[AR(1) models explained]
.blockquote[
$$y_{t}=c + \phi_1 y_{t -1}+\varepsilon_{t}$$
* When $\phi_1=0$, $y_t$ is **equivalent to White Noise**
* When $\phi_1=1$ and $c=0$, $y_t$ is **equivalent to a Random Walk**
* When $\phi_1=1$ and $c\ne0$, $y_t$ is **equivalent to a Random Walk with drift**
* When $\phi_1<0$, $y_t$ tends to **oscillate between positive and negative values**.
]
]
]

---
class: middle

## Moving Average (MA) models

.blockquote.midi[

Moving Average (MA) models:
$$y_{t} =c +\varepsilon_t + \theta_{1}\varepsilon_{t - 1} + \theta_{2}\varepsilon_{t - 2} +\cdots+\theta_{q}\varepsilon_{t - q},$$
- where $\varepsilon_t$ is white noise.
- This is a multiple regression with  **past errors** as predictors. **Don't confuse this with moving average smoothing!**
]

```{r maq}
set.seed(2)
p1<-autoplot(20 + arima.sim(list(ma = 0.8), n = 100)) +
  ylab("") + ggtitle(TeX(r'(MA(2) model: $y_t = 20 + \epsilon_t + 0.8 \epsilon_{t-1} ... \epsilon_t \sim N(0,1)$)'))
p2 <- autoplot(arima.sim(list(ma = c(-1, +0.8)), n = 100)) +
  ylab("") + ggtitle(TeX(r"(MA(2) model:  $y_t = \epsilon_t -\epsilon_{t-1} + 0.8 \epsilon_{t-2} ... \epsilon_t \sim N(0,1)$)"))
gridExtra::grid.arrange(p1,p2,nrow=1)
```

---
class: middle

.hand-large[Putting it all together]

### ARIMA models

- Autoregressive Integrated Moving Average models

.blockquote[
Autoregressive Moving Average models
$$
\begin{align*}
y_{t} = c+ \phi_{1}y_{t - 1} +\cdots  +\phi_{p}y_{t-p} \\
 \theta_{1}\varepsilon_{t - 1} + \cdots +\theta_{q}\varepsilon_{t-q} +\varepsilon_{t}.
\end{align*}
$$
]

--

* Predictors include both **lagged values of $y_t$ and lagged errors.**
* Conditions on coefficients ensure stationarity.
* Conditions on coefficients ensure invertibility.
* Combine ARMA model with **differencing**.

---
class: middle 

## ARIMA models notation

.pull-left[
>Autoregressive Integrated Moving Average models
- ARIMA(p, d, q) model
- AR part: p = order of the autoregressive part 
- I part: d = degree of first differencing involved
- MA part: q = order of the moving average part.

- ARIMA(1,1,1) model:
$$(y_t-y_{t-1}) =c + \phi_1 (y_{t-1}- y_{t-2}) +\theta_1\varepsilon_{t-1} + \varepsilon_t$$

]
.pull-right[
* White noise model:  ARIMA(0,0,0)
* Random walk:  ARIMA(0,1,0) with no constant
* Random walk with drift:  ARIMA(0,1,0) with `constant term`
* AR(p): ARIMA(p,0,0)
* MA(q): ARIMA(0,0,q)
]


---
class: middle

## ARIMA modelling of US consumption

.panelset[
.panel[
.panel-name[The data]
```{r}
autoplot(uschange[,"Consumption"]) +
  xlab("Year") +
  ylab("Quarterly percentage change") +
  ggtitle("US consumption")
```
]
.panel[
.panel-name[fit an ARIMA(2,0,2)]
```{r, echo=TRUE}
((fit <- arima(uschange[,"Consumption"],order = c(2,0,2))))
```

```{r usconsumptioncoefs, echo=FALSE}
coef <- coefficients(fit)
ar1=round(coef['ar1'],3) %>% unname()
ar2=round(coef['ar2'],3) %>% unname()
intercept = round(coef['intercept'] * (1-coef['ar1'] - coef['ar2']),3) %>% unname()
ma1=round(coef['ma1'],3) %>% unname()
ma2=round(coef['ma2'],3) %>% unname()
sigma=round(sqrt(fit$sigma2),3) %>% unname()
sigma2=round(fit$sigma2,3) %>% unname()
```
]
.panel[
.panel-name[Model estimates in math]

- $y_t = c + `r ar1`y_{t-1} `r ar2`y_{t-2}`r ma1` \varepsilon_{t-1}+ `r ma2`\varepsilon_{t-2}+ \varepsilon_{t}$
- where $c= `r intercept`$
- and $\varepsilon_t$ is white noise with a standard deviation of $`r sigma` = \sqrt{`r sigma2`}$. 

]
.panel[
.panel-name[Forecasts]

```{r, echo=TRUE}
fit %>% forecast(h=10) %>% autoplot(include=80)
```
]
]

---
class: middle

## Understanding ARIMA models

.large[
* If $c=0$ and $d=0$, the long-term forecasts will go to zero.
* If $c=0$ and $d=1$, the long-term forecasts will go to a non-zero constant.
* If $c=0$ and $d=2$, the long-term forecasts will follow a straight line.
* If $c\ne0$ and $d=0$, the long-term forecasts will go to the mean of the data.
* If $c\ne0$ and $d=1$, the long-term forecasts will follow a straight line.
* If $c\ne0$ and $d=2$, the long-term forecasts will follow a quadratic trend.
]
---
class: middle

## Understanding ARIMA models
### Forecast variance and $d$

* The higher the value of $d$, the more rapidly the prediction intervals increase in size.
* For $d=0$, the long-term forecast standard deviation will go to the standard deviation of the historical data.

### Cyclic behaviour
* For cyclic forecasts,  $p\ge2$ and some restrictions on coefficients are required.
* If $p=2$, we need $\phi_1^2+4\phi_2<0$. Then average length of stochastic cycles is

$$(2\pi)/\left[\text{arc cos}(-\phi_1(1-\phi_2)/(4\phi_2))\right].$$

* This formula has important uses in estimation business and economic cycles. (See Example 2.3 in Tsay (2010))

---
class: middle

# Estimation and order selection

## Maximum likelihood estimation

- Having identified the model order, we need to estimate the parameters $c,\phi_1,\dots,\phi_p \text{  }\theta_1,\dots,\theta_q$.

* MLE is very similar to least squares estimation obtained by minimizing $\sum_{t-1}^T e_t^2$
* The `Arima()` command allows CLS or MLE estimation.
* Non-linear optimization must be used in either case.
* Different software will give different estimates.

---
class: middle

## Partial autocorrelations

.blockquote[
- Partial autocorrelations} measure relationship between $y_{t}$  and  $y_{t - k}$, when the effects of other time lags $1,2, 3, \dots, k - 1$are removed.

- $\alpha_k$= $k$th partial autocorrelation coefficient
- $\alpha_k${equal to the estimate of $b_k$ in regression:

$$y_t = c + \phi_1 y_{t-1} + \phi_2 y_{t-2} + \dots + \phi_k y_{t-k}$$

* Varying number of terms on RHS gives $\alpha_k$ for different values of $k$.
* There are more efficient ways of calculating $\alpha_k$.
* $\alpha_1=\rho_1$
* same critical values of $\pm 1.96/\sqrt{T}$ as for ACF.
]

---
class: middle

## Example: US consumption

```{r usconsumptionacf}
p1 <- ggAcf(uschange[,"Consumption"],main="")
p2 <- ggPacf(uschange[,"Consumption"],main="")
gridExtra::grid.arrange(p1,p2,nrow=1)
```

---
class: middle

## ACF and PACF interpretation

**AR(1)**
$$rho_k =\phi_1^k \text{  for k=1,2,}\dots$$

$$\alpha_1= \phi_1 \alpha_k = 0\text{for k=2,3}\dots$$

So we have an AR(1) model when

* autocorrelations exponentially decay
* there is a single significant partial autocorrelation.

---
class: middle

## ACF and PACF interpretation
.three-column[
**AR(p)**
  * ACF dies out in an exponential or damped sine-wave manner
  * PACF has all zero spikes beyond the $p$th spike

So we have an AR(p)) model when

  * the ACF is  exponentially decaying or sinusoidal
  * there is a significant spike at lag $p$ in PACF, but none beyond $p$
]

.three-column[
**MA(1)**
$$
\begin{align*}
\rho_1 &= \theta_1 \qquad \rho_k = 0\qquad\text{for k=2,3,...};\\
\alpha_k &= -(-\theta_1)^k
\end{align*}
$$
So we have an MA(1) model when
 * the PACF is  exponentially decaying and
 * there is a single significant spike in ACF
]
.three-column[
**MA(q)**

 * PACF dies out in an exponential or damped sine-wave manner
 * ACF has all zero spikes beyond the $q$th spike

So we have an MA(q) model when

  * the PACF is  exponentially decaying or sinusoidal
  * there is a significant spike at lag $q$ in ACF, but none beyond $q$
]

---
class: middle

## Information criteria

.blockquote[
**Akaike's Information Criterion (AIC):**

$\text{AIC} = -2 \log(L) + 2(p+q+k+1),$  where $L$ is the likelihood of the data, $k=1$ if $c\ne0$ and $k=0$ if $c=0$]

.blockquote[
**Corrected AIC:** <br>
$\text{AICc} = \text{AIC} + \frac{2(p+q+k+1)(p+q+k+2)}{T-p-q-k-2}.$
]

.blockquote[
**Bayesian Information Criterion:**<br>
$\text{BIC} = \text{AIC} + [\log(T)-2](p+q+k-1).$<br>
Good models are obtained by minimizing either the AIC, AICc or BIC. My preference is to use the AICc.]


---
class: middle

# Powerful non-stationary model in finance

- In financial time series an important class on non-stationary times series model is the random walk model
- A random walk can be define as $y_t=y_{t-1}+ error_t$ or its drift variation $y_t= constant + y_{t-1}+ error_t$

.panelset[
.panel[
.panel-name[Simulation 1]
.pull-left-narrow[
$y_t = 10 + 0.99y_{t-1}+ \varepsilon_t$
```{r rw_drift1, eval=FALSE}
set.seed(1)
autoplot(10 + arima.sim(list(ar =0.99), n = 100))  + ylab("") + 
  ggtitle("Is this a random walk with drift?")
```
]
.pull-right-wide[
```{r rw_drift2, echo=FALSE}
set.seed(1)
autoplot(10 + arima.sim(list(ar =0.99), n = 100))  + ylab("") + 
  ggtitle("Is this a random walk with drift?")
```
]
]
.panel[
.panel-name[Simulation 1]
.pull-left-narrow[
```{r rw_drift3,eval=FALSE}
set.seed(2)
S0=10
n=100
chgs=rnorm(n-1,1.001,0.01)
rw=ts(cumprod(c(S0,chgs)))
autoplot(rw)  + ylab("") + 
  ggtitle("Is this a random walk with drift?")
```
]
.pull-right-wide[
```{r rw_drift4,echo=FALSE}
set.seed(2)
S0=10
n=100
chgs=rnorm(n-1,1.001,0.01)
rw=ts(cumprod(c(S0,chgs)))
autoplot(rw)  + ylab("") + 
  ggtitle("Is this a random walk with drift?")
```

]
]
]

---
class: middle

# A rule-based ARIMA algorithm

.blockquote[`auto.arima()`]

- For a non-seasonal ARIMA process we first need to select appropriate orders: $p,q,d$
- We use the [Hyndman and Khandakar (JSS, 2008)](https://www.jstatsoft.org/article/view/v027i03) algorithm:

.blockquote[
* Select no. differences $d$ and $D$ via KPSS test and seasonal strength measure.
* Select $p,q$ by minimising AICc.
* Use stepwise search to traverse model space.
]

---
class: middle

## How does auto.arima() work?

.blockquote[
$$\text{AICc} = -2 log(L) + 2(p+q+k+1)\left[1 +
\frac{(p+q+k+2)}{T-p-q-k-2}\right].$$

- where $L$ is the maximised likelihood fitted to the differenced
data,

- $k=1$ if $c\neq 0$ and $k=0$ otherwise.
]

---
class: middle

# Algorithmic rule 1:

- Select current model (with smallest AICc) from:
- ARIMA$(2,d,2)$
- ARIMA$(0,d,0)$
- ARIMA$(1,d,0)$
- ARIMA$(0,d,1)$

---
class: middle

# Algorithmic rule 2:

- Consider variations of current model:

* vary one of $p,q,$ from current model by $\pm1$;
* $p,q$ both vary from current model by $\pm1$;
* Include/exclude $c$ from current model.

.heatinline[
Model with lowest AICc becomes current model.
]

.large[Repeat rule 2 until no lower AICc can be found]

---
class:middle

## Example: VIX index

.panelset[
.panel[.panel-name[Human choice]

.pull-left[
```{r, echo=T,eval=F}
tidyquant::tq_get('^VIX') %>%
  select(adjusted) %>%
  ts() -> vix_ts
ggtsdisplay(log(vix_ts))
```
]
.pull-right[
```{r}
tidyquant::tq_get('^VIX') %>%
  select(adjusted) %>%
  ts() -> vix_ts
ggtsdisplay(log(vix_ts))
```
]
]
.panel[.panel-name[Human choice]

.pull-left[
```{r, eval=F,echo=T}
ggtsdisplay(diff(log(vix_ts)))
```

]
.pull-right[
```{r}
ggtsdisplay(diff(log(vix_ts)))
```
]
]
.panel[.panel-name[Human choice]

.pull-left[
```{r, eval=F,echo=T}
(fit <- Arima(log(vix_ts),order=c(7,1,0)))
```

]
.pull-right[
```{r}
(fit <- Arima(log(vix_ts),order=c(7,1,0)))
```
]
]
.panel[.panel-name[Algorithmic choice]

.pull-left[
```{r,eval=F,echo=TRUE}
auto.arima(log(vix_ts))
```

]
.pull-right[
```{r}
auto.arima(log(vix_ts))
```
]
]
.panel[.panel-name[Try harder algorithm]

.pull-left[
```{r tryharder,eval=F,echo=TRUE}
(fit.auto<-auto.arima(log(vix_ts), 
                      stepwise=FALSE,
                      approximation=FALSE))
```

]
.pull-right[
```{r}
(fit.auto<-auto.arima(log(vix_ts), 
                      stepwise=FALSE,
                      approximation=FALSE))
```
]
]
.panel[.panel-name[Discussion]
.discussion[
* Setting both `stepwise` and `approximation` arguments to `FALSE` will slow the automation down but provides a more exhaustive search for the appropriate model.
* The `auto.arima` function then searches over all possible models using MLE.
* See `help(auto.arima)` for more details.
]
]
]

---
class: middle

## Human Vs Algo: Residual Diagnostics

.panelset[
.panel[.panel-name[Human choice]

.pull-left[
```{r,eval=F,echo=T}
checkresiduals(fit, test=FALSE)
```

]
.pull-right[
```{r}
checkresiduals(fit, test=FALSE)
```

]
]

.panel[.panel-name[Human forecasting]
.pull-left[
```{r, eval=F,echo=T}
fit %>% forecast(h=252) %>% autoplot
```
]
.pull-right[
```{r}
fit %>% forecast(h=252) %>% autoplot
```
]
]

.panel[.panel-name[Algorithm]

.pull-left[
```{r,eval=F,echo=T}
checkresiduals(fit.auto,test = FALSE)
```
]
.pull-right[
```{r}
checkresiduals(fit.auto,test = FALSE)
```
]
]
.panel[.panel-name[Algo forecasting]

.pull-right[
```{r,echo=T,eval=F}
fit.auto %>% forecast(h=252) %>% autoplot()
```
]
.pull-right[
```{r}
fit.auto %>% forecast(h=252) %>% autoplot()
```
]
]
]

---
class: middle

## Modelling procedure with `Arima`

.blockquote[
1. Plot the data. Identify any unusual observations.
2. If necessary, transform the data (using a Box-Cox transformation) to stabilize the variance.
3. If the data are non-stationary: take first differences of the data until the data are stationary.
4. Examine the ACF/PACF: Is an AR(p) or MA(q) model appropriate?
5. Try your chosen model(s), and  use the AICc to search for a better model.
6. Check the residuals from your chosen model by plotting the ACF of the residuals, and doing a portmanteau test of the residuals. If they do not look like white noise, try a modified model.
7. Once the residuals look like white noise, calculate forecasts.
]

---
class: middle

## Modelling procedure with `auto.arima`

1. Plot the data. Identify any unusual observations.
2. If necessary, transform the data (using logs) to stabilize the variance.
3. Use `auto.arima` to select a model.

6. Check the residuals from your chosen model by plotting the ACF of the residuals, and doing a portmanteau test of the residuals. If they do not look like white noise, try a modified model.
7. Once the residuals look like white noise, calculate forecasts.

---
class: middle

# Project based example

.hand[Russell 2000 index monthly returns]

.pull-left[
```{r r2000r1,echo=TRUE}
tsfe::indices %>%
  select(date,`RUSSELL 2000 - PRICE INDEX`) %>%
  rename(r2000=`RUSSELL 2000 - PRICE INDEX`) %>%
  drop_na() %>%
  tq_transmute(select =r2000,mutate_fun = periodReturn,type='log') ->monthly_r2002r
  ts(monthly_r2002r$monthly.returns, start = c(1988,1))->r2000r_m_ts
autoplot(r2000r_m_ts) + xlab("Year") +
  ylab("monthly log returns")
```
]
.pull-right[
1. Time plot shows sudden changes, particularly big movements in 2007/2008 due to financial crisis. Otherwise nothing unusual and no need for  data adjustments.
2. Little evidence of changing variance, so no log transformation needed.
3. Data are clearly stationary, so no *differencing* required.
]

---
class: middle

# Project based example

.hand[Russell 2000 index monthly returns]

.pull-left[

```{r r2000r2,echo=TRUE}
ggtsdisplay(r2000r_m_ts)
```
]
.pull-right[
4. PACF is suggestive of AR(5). So initial candidate model is ARIMA(5,0,0). No other obvious candidates.
5. Fit ARIMA(5,0,0) model along with variations: ARIMA(4,0,0), ARIMA(3,0,0), ARIMA(4,0,1), etc. ARIMA(3,0,1) has smallest *AICc* value.
]

---
class: middle

# Project based example

.panelset[
.panel[.panel-name[Manual fit]
.pull-left[
```{r,echo=T,eval=F}
(fit <- Arima(r2000r_m_ts, order=c(3,0,1)))
```
]
.pull-right[
```{r r2000r3}
(fit <- Arima(r2000r_m_ts, order=c(3,0,1)))
```
]
]
.panel[.panel-name[Residual check]
6. ACF plot of residuals from ARIMA(3,0,1) model look like white noise.

```{r r20005,echo=T}
checkresiduals(fit, plot=FALSE)
```
]
.panel[.panel-name[Forecasting]
```{r,echo=T}
fit %>% forecast(h=24) %>% autoplot
```
]
]

---
class: middle

## Understanding the ARIMA output

```{r r2000r6}
(fit <- Arima(r2000r_m_ts, order=c(3,0,1)))
```

---
class: middle

## Understanding the ARIMA output

```{r coeffs}
a=unname(round(fit$coef["intercept"],3))
ar1=unname(round(fit$coef["ar1"],3))
ar2=unname(round(fit$coef["ar2"],3))
ar3=unname(round(fit$coef["ar3"],3))
ma1=unname(round(fit$coef["ma1"],3))
```

* The fitted model is:

$$y_t=`r ar1`y_{t-1}`r ar2`y_{t-2}`r ar3`y_{t-3} +`r ma1`\varepsilon_{t-1}+0.008$$

* The standard errors are 0.13, 0.06, 0.05, 0.12 and 0.002, respectively.
* This suggest that only the AR1 and the constant (mean) are more than 2 SEs away from zero and thus statistically significant.
* The significance of $\phi_0$ of this entertained model implies that the expected mean return of the series is positive.

* In fact $\hat{\mu}=`r a`/(1-(`r ar1``r ar2``r ar3`)) =`r round(a/(1-sum(ar1,ar2,ar3)),4)`$ which is small but has long term implications.

* Using the multi-period return definition from the financial data lecture an annualised log return is simple $\sum_1^{12} y_t$ $\approx `r round(12*a/(1-sum(ar1,ar2,ar3)),2)`$ per annum. 

---
class: middle

## Frequentist prediction intervals

$$\hat{y}_{T+h|T} \pm 1.96\sqrt{v_{T+h|T}}$$
where $v_{T+h|T}$ is estimated forecast variance.

--

* $v_{T+1|T}=\hat{\sigma}^2$ for all ARIMA models regardless of parameters and orders.
* Multi-step prediction intervals for ARIMA(0,0,$q$):

$$\displaystyle y_t = \varepsilon_t + \sum_{i=1}^q \theta_i \varepsilon_{t-i}$$

$$\displaystyle v_{T|T+h} = \hat{\sigma}^2 \left[ 1 + \sum_{i=1}^{h-1} \theta_i^2\right], \qquad\text{for~} h=2,3,\dots.$$

* AR(1): Rewrite as MA($\infty$) and use above result.
* Other models beyond scope of this subject.

---
class: middle

## Prediction intervals

* Prediction intervals **increase in size with forecast horizon**.
* Prediction intervals can be difficult to calculate by hand
* Calculations assume residuals are **uncorrelated** and **normally distributed**.
* Prediction intervals tend to be too narrow.
    * the uncertainty in the parameter estimates has not been accounted for.
    * the ARIMA model assumes historical patterns will not change during the forecast period.
    * the ARIMA model assumes uncorrelated future $errors$


